---
title: "3. User Defined R Functions"
author: "Cheng Peng"
date: "Lab Note for MAT325 Numerical Analysis "
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    fig_width: 6
    fig_caption: yes
    number_sections: yes
    theme: readable
    fig_height: 6
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 5
    fig_height: 5
editor_options: 
  chunk_output_type: inline
---

```{=html}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: left;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```
```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("nleqslv")) {
   install.packages("nleqslv")
   library(nleqslv)
}
#
# specifications of outputs of code in code chunks
knitr::opts_chunk$set(echo = TRUE,      # include code chunk in the output file
                      warnings = FALSE,  # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      messages = FALSE,  #
                      results = TRUE     # you can also decide whether to include the output
                                         # in the output file.
                      )   
```

\


# Introduction

One of the great strengths of R is that it allows users to extend the capacity of R through user-defined functions. Functions are often used to encapsulate a sequence of expressions that need to be executed repetitively. We have made code to implement bisection and fixed-point methods by specific examples. We may want to write the R function of these implementations so we can use them for other root-finding problems with the appropriate input information. 


# Syntax and Types of User-Defined R Functions

The syntax of the general user-defined R function has the following form

```
myfunction <- function(arg1, arg2, ... ){
   statements
   return(object)
}
```

`arg1, arg2, ...` are arguments that are passed into the function. The arguments could be any objects such as vectors and other user-defined R functions.



**Example 1** Finding the standard deviation of an input vector.

```{r}
sdfun <- function(x) {
  res <- sqrt(sum((x - mean(x))^2) / (length(x) - 1))
  return(res)
}
##
vec = c(1,4,2,6,-3, -5)
sdfun(x = vec)   # or simply sdfun(vec)
```

We have defined functions in implementing root-finding methods.

\

# Writing R Functions for Root-finding Methods

As an example, we will write an R function to implement the bisection method for finding the root for any given equation on ver an interval (if it exists). Two versions of R functions will be presented in the following.

## Function with Numerical Outputs

We simply wrap up the example code in the lecture note to make the following function.

```{r}
num.FixedPoint = function(fn,       # function that satisfies f(x) = x
                          a,        # lower limit of the interval [a, b]
                          b,        # upper limit of the interval [a, b]
                          TOL,      # error tolerance
                          N,        # maximum number of iterations
                          x0,       # initial value of x
                          detail,    # intermediate output
                          ...){
 gfun = fn
 x = x0
 ERR = Inf
 n = 0
## 
 while (ERR > TOL){
  n = n + 1
  new.x = gfun(x)
  ERR = abs(new.x - x)
  if(ERR < TOL){
    cat("\n\nThe algorithm converges!")
    cat("\nThe approximate root is:", new.x,".")
    cat("\nThe absolute error is:", ERR, ".")
    cat("\nThe number of iterations is:", n, ".")
    break
  } else{
    if(ERR > 10^7){
        cat("\n\nThe algorithm diverges!")
        break
    } else{
         if(detail == TRUE){
            cat("\nIteration:",n,". Estimated root:", new.x, ". Absolute error:", ERR,".")
          }
         x = new.x         # update x value!!!
    }
  }
  if(n == N){
    cat("\n\nThe maximum number of iterations is achieved!")
    break
  } 
 }
}

```

Next, we use several examples.

**Example 1** Using the fixed-point method to find the approximate root of $x^3 - 7x +2 = 0$. To use the fixed-point method, we rewrite the equation into the form $(x^3 + 2)/7 = x$. Then $g(x) = (x^3 + 2)/7$ will be the function to be passed into the function.

```{r}
###
fun0 = function(x) (x^3 + 2)/7
num.FixedPoint(fn = fun0, a = 0, b = 2, TOL = 10^(-6), N = 200, x0 =1.5, detail = TRUE)
```

**Example 2**: Calculate $1/\sqrt{2}$ by using the fixed-point method. Note that $f(x) = x^2 - 1/2$. To use the fixed point method, we need $g(x) = x -x^2 +1/2$ as the input function.

```{r}
###
fun0 = function(x) x - x^3 + 1/2
num.FixedPoint(fn = fun0, a = -1, b = 2, TOL = 10^(-5), N = 200, x0 =0.7, detail =FALSE)
```

## Function with More Optional Outputs

We include graphics the function created previously.

```{r fig.align='center', fig.height=6, fig.width=6}
FixedPointAlgR = function(fn,       # function that satisfies f(x) = x
                          a,        # lower limit of the interval [a, b]
                          b,        # upper limit of the interval [a, b]
                          TOL,      # error tolerance
                          N,        # maximum number of iterations
                          x0,       # initial value of x
                          detail=TRUE,   # intermediate numerical outputs
                          graphic=TRUE,  # intermediate graphic outputs
                          x.lim,    # x-axis of graphic window
                          y.lim,    # y-axis of graphic window
                          sleep,    # sleep time between iterations
                          ...){
 gfun = fn      # rename the input function used in the function
 x = x0         # rename the initial value used in the function
 ERR = Inf      # initial error (a big  number to start the iteration)
 n = 0          # initial iteration counter - iterator or index
##
 if(graphic==TRUE){
 xlimit=c(a-0.1*abs(b-a), b+0.1*abs(b-a))               # x-limit in the graphic window
 xx=seq(a-0.1*abs(b-a), b+0.1*abs(b-a), length = 2000)  # sequence of x values
 yy1 = gfun(xx)                                         # evaluate g(x)
 yy2 = xx                                               # evaluate diagonal
 plot(xx, yy1, ylim = y.lim, xlim = x.lim, type = "l", lwd = 2, lty = 1, col = "blue")
 lines(xx,yy2, lwd = 2, lty = 1, col ="darkred")        # add line y = x
 title("Fixed Point Algorithm Approximation")           # add title to the graphic
 }
## 
 while (ERR > TOL){
  Sys.sleep(sleep)                                      # put the system to sleep
  n = n + 1
  new.x = gfun(x)
  ERR = abs(new.x - x)
  if(ERR < TOL){
    cat("\n\nThe algorithm converges!")
    cat("\nThe approximate root is:", new.x,".")
    cat("\nThe absolute error is:", ERR, ".")
    cat("\nThe number of iterations is:", n, ".")
    break
  } else{
    if(ERR > 10^7){
        cat("\n\nThe algorithm diverges!")
        break
    } else{
         if(detail == TRUE){
            cat("\nIteration:",n,". Estimated root:", new.x, ". Absolute error:", ERR,".")
         }
        if(graphic==TRUE){
            segments(new.x, new.x,       new.x,       gfun(new.x), col="purple")
            segments(new.x, gfun(new.x), gfun(new.x), gfun(new.x), col="purple")
           }

         x = new.x         # update x value!!!
    }
  }
  if(n == N){              # checking iteration limit
    cat("\n\nThe maximum number of iterations is achieved!")   
    break
  } 
 }
}

```

\

**Example 1** (Revisited)

```{r}
###
fun0 = function(x) (x^3 + 2)/7
FixedPointAlgR(fn = fun0, a = 0, b = 1, TOL = 10^(-6), N = 200, 
               x0 =0.35, detail = TRUE, graphic = FALSE, 
               x.lim=c(0.285, 0.294), y.lim=c(0.28, 0.32), sleep = 0.1)
```

\

**Example 2** (Revisited)

```{r}
###
fun0 = function(x) x - x^3 + 1/2
FixedPointAlgR(fn = fun0, a = -1, b = 2, TOL = 10^(-5), N = 200, 
               x0 =-1.2, detail =FALSE, graphic = TRUE, 
               x.lim=c(0.7, 0.9), y.lim=c(0.6, 0.9), sleep = 0.1)
```










