---
title: "Midterm Exam One"
author: "MAT325 Numerical Analysis"
date: "3/10/2023"
output:
  html_document: 
    toc: no
    toc_depth: 4
    toc_float: yes
    fig_width: 4
    fig_caption: yes
    number_sections: no
    theme: readable
    fig_height: 4
  word_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    number_sections: no
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{=html}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the 
presentation of a document written in HTML or XML. it is a simple mechanism for 
adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: center;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```
```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("nleqslv")) {
   install.packages("nleqslv")
   library(nleqslv)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
#
options(digits = 15)
# specifications of outputs of code in code chunks
knitr::opts_chunk$set(echo = TRUE,      # include code chunk in the output file
                      warnings = FALSE,  # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      messages = FALSE,  #
                      results = TRUE     # you can also decide whether to include the output
                                         # in the output file.
                      )   
```


\

\

# Instructions

\

**General Information**: This is an open book and open note exam. You can use the textbook, lecture notes, R code/function in lecture/lab notes, and internet resources for the exam. However, you must complete the exam independently. No collaboration is allowed.

**Specific Requirements**:

1. Use RMarkdown or other typesetting software programs to prepare your solution.

2. The code must be commented on and included in the solution. If you use a typesetting program other than RMarkdown, please put the code in the appendix. 

3. Please upload a copy of your solution in PDF format and a copy source file to the D2L drop box.

4. Please feel free to use/modify the code from the lectures and those you developed in the assignments.

\

\

# Problem Set


## Problem 1: Finding the root of a non-linear equation

Consider the function $f(x) = \tan(\sin(x^3))$ over interval $[0, \sqrt{\frac{5}{2}}]$. 


1. Choose 500 equally-spaced base points in the interval $[0, \sqrt{\frac{5}{2}}]$ and plot the curve of the function.


2. Find all roots of equation $\tan(\sin x^3) = 0$ mathematically over interval $\left[0, \sqrt{\frac{5}{2}}\right]$.


3. Use the Newton method to find these roots (use $TOL = 10^{-6}$) and compare these approximate roots with their corresponding true roots.


\


## Problem 2. Approximation with interpolated Polynomials


The Laguerre polynomial that has applications in differential equations and physics (particularly in quantum mechanics) that is  is defined by

$$
L_n(x) = \sum_{k=0}^n {n \choose k}\frac{(-1)^k}{k!}x^k
$$

In this problem, we only consider the degree 4 Laguerre polynomial 

$$
L_4(x) = \frac{x^4}{24} - \frac{2x^3}{3} + 3x^2 - 4x + 1
$$

over interval $[0,10]$.

```{r fig.align='center', fig.height=4, fig.width=4}
xx= seq(0, 10, length = 200)
yy = xx^4/24 - (2/3)*xx^3 + 3*xx^2 - 4*xx + 1
plot(xx, yy, type = "l", lwd = 2, col = "navy",
     xlab = "", ylab = "",
     main = "Degree 4 Laguerre Polynomial")
abline(v=c(0, 10), h = 0, lty = 2, col = "blue")
```

The purpose of this problem is to assess the performance of the following three different approximated polynomials:


1. The degree 4 Taylor polynomial expanded at $x = 7.75$.


2. The Newton interpolated polynomial using unequally-spaced nodes $x =0.3 , 0.95, 3.3,  7.75, 9.4$


3. The Newton interpolated polynomial using equally-spaced nodes $x = 1, 3, 5, 7, 9$


Evaluate the above three approximated polynomials and the original Laguerre polynomial at $x = 3$ and compare the three approximated values with the true value obtained from the original Laguerre polynomial. Which polynomial gives the best approximation? 



\


## Problem 3. Application - Approximating Unknown Function and Making Prediction 

The table below gives the actual thermal conductivity data for the element mercury. The objective is to find the analytic expression that approximates the relationship between temperature and pressure.


|Temperature (Fahrenheit), T | 220  |  230 |  240 |  250 |  260 |  270 | 280  | 290  | 300  |
|:----------------------------------|:-----|:-----|:-----|:-----|:-----|:-----|:-----|:-----|
|Pressure (Pound), P         |17.19 |20.78 |24.97 |29.82 |35.42 |41.85 |49.18 |57.53 |66.98 |

Use the Newton interpolation and all given data points to construct a polynomial of degree 8 and use this interpolating polynomial to **predict** the corresponding pressures for temperatures $T = 235, 255, 279, 295$. 

